\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-16 NAR_DRAFT.tex
add calibration plots
add wgs figure
add figures

\subsection{The likelihood function}
We designed \batcave to be used as a post-calling evaluation tool with MuTect
MuTect is particularly useful because it directly reports the log ratio of the of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

\section{RESULTS}
We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Receiver operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.}
\label{NAR-wes_fig}
\end{figure*}
\subsection{Convergence of the site-specific prior to the data generating distribution}
The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for our real datasets are \bkmcomment{X} and for our simulated data sets are 1500 to 5000.

@@

@@

We measured calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
This is useful in our setting because the density of predicted probabilities tends to be strongly bi-modal.
As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (top density curves), and decrease the probability of low probability but false variants (bottom density curves.).
For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse.
This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.

@@

\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wes.pdf}
  \end{center}
  \caption{Model performance on 100X simulated whole genome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
\label{NAR-wgs_fig}
\end{figure*}

\begin{table*}[b]
  \tableparts{%
  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
  \label{table:01}%
  }{%
  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
  \toprule
  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  \citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular*}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
  \end{table*}
\end{document}

>>>>

\subsection{The likelihood function}
We designed \batcave to be used as a post-calling evaluation tool with MuTect.
MuTect is particularly useful because it directly reports the log ratio of the likelihood functions for the null and alterative hypotheses in \ref{eqn:hypothesis} as either \texttt{TLOD} (MuTect1) or \texttt{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all of the analysis in this paper, so the output of MuTect provides us with

@@

\section{RESULTS}
We designed \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
The algorithm is effecient, processing 22,000 variants per second on a typical iMac desktop computer, which corresponds to 100 seconds to process a 500X exome and 2000 seconds for a 100X whole genome.
MuTect is widely used, has state of the art sensitivity and specificity, and includes numerous heuristic filters and alignment adjustments that reduce the prevalence of sequencing errors in results \cite{Griffith2015}.
\batcave takes the output from MuTect, extracts the log-likelihood ratio for each potential variant site, and uses that ratio to separate the potential sites in to high-confidence and low-confidence groups.
The mutation profile is computed from the high confidence sites, and the posterior probability of mutation is recomputed for all sites as described in Materials and Methods.
\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Receiver operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.}
\label{NAR-wes_fig}
\end{figure*}

\subsection{Convergence of the site-specific prior to the data generating distribution}
The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure \ref{NAR-sigfig} \textbf{e} and \textbf{f} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points.}
\label{NAR-ppv_fig}
\end{figure}

@@

We measured overall calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
A large fraction of true negative variants have posterior probabilities of less than $10{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse.
This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.

@@

\begin{figure*}
  \begin{center}
  \includegraphics{figures/fig_wgs.pdf}
  \end{center}
  \caption{Model performance on 100X simulated whole genome.
  \textbf{(a)} Precision recall curves and \textbf{(c)} Reciever operating characteristic curves for 3 mutation signatures.
  Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{b)} the MuTect model and \textbf{d)} the \batcave model.
  A perfectly calibrated model would generate the diagonal line.\bkmcomment{change the axis on b and d to be the same as the steyerberg paper. much clearer}}
\label{NAR-wgs_fig}
\end{figure*}

\begin{table*}[b]
  \tableparts{%
  \caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
  \label{table:01}%
  }{%
  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
  \toprule
  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & -- & --
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & -- & --
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & -- & --
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  \citet{Williams2016} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular*}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
  \end{table*}

\begin{figure}
  \begin{center}
  \includegraphics{figures/ppv_wgs.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points.}
\label{NAR-ppv_wgs_fig}
\end{figure}

\end{document}
